---
title: "GII Index Reduction Analysis Report"
author: "JRC"
date: "`r Sys.Date()`"
format:
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
bibliography: references.ris
header-includes:
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
source("scripts/R/config.R")
source("scripts/R/functions.R")
```

```{r load-structure, include=FALSE}
# Structural facts about the dataset/index - computed from source metadata and
# already-processed data files, never hand-typed.
imeta <- readxl::read_excel("gii-data/imeta.xlsx")
  # strip stray annotation marks (e.g. "...business†", "...tariff rate*")
  imeta$iName <- gsub("[*†]", "", imeta$iName)
  imeta$iName <- trimws(imeta$iName)
  imeta$iName <- trimws(gsub("\\s+", " ", imeta$iName)) # collapse repeated spaces, trim ends
  
standardized_data <- read.csv("data/gii_data_standardized.csv")

n_countries <- nrow(standardized_data)
n_indicators_full <- ncol(standardized_data) - 1  # exclude Country column
n_subindices <- sum(imeta$Level == 4)
n_index <- sum(imeta$Level == 5)

pillar_sizes <- group_indicator_counts(CONFIG$pillars_level3, imeta, level = 3)
subpillar_sizes <- group_indicator_counts(CONFIG$subpillars_level2, imeta, level = 2)
pillar_names <- imeta |> filter(Level == 3) |> select(NUM, iName) |> tibble::deframe()
subpillar_names <- imeta |> filter(Level == 2) |> select(iCode, iName) |> tibble::deframe()
indicator_names <- imeta |> filter(Level == 1) |> select(iCode, iName) |> tibble::deframe()

# Map a character vector of indicator iCodes to their full names, e.g. for
# the results of a set operation (intersect/setdiff) on selected-indicator sets.
name_indicators_vec <- function(codes) unname(indicator_names[codes])
```

```{r load-results, include=FALSE}
# All analysis-result files this report references, loaded once up front so
# every number quoted below traces back to a file produced by the pipeline
# scripts rather than being typed by hand.
allocations <- readRDS(file.path("outputs", "allocation20.rds")) |>
  mutate(combinations = choose(n_indicators, allocation))

fig2_results <- read.csv("outputs/fig2_rank_correlations.csv")
strategy1_country <- read.csv("outputs/exhaustive_strategy1_pillar.csv")
strategy2_country <- read.csv("outputs/exhaustive_strategy2_subpillar.csv")
rand_index_flat <- read.csv("outputs/clustering_rand_index.csv")
strategy1b <- read.csv("outputs/exhaustive_strategy1b_pillar.csv")
strategy1a <- read.csv("outputs/exhaustive_strategy1a_pillar.csv")
strategy2a <- read.csv("outputs/exhaustive_strategy2a_subpillar.csv")
rand_index_proportional <- read.csv("outputs/clustering_rand_index_proportional.csv")

# Human-readable labels for the allocation rules. The internal codes are never
# shown to the reader - every table and every sentence uses these names.
strategy_labels <- c(
  Strategy1_3perPillar             = "Flat allocation (pillar)",
  Strategy2_1perSubpillar          = "Flat allocation (sub-pillar)",
  Strategy1a_ProportionalPillar    = "Proportional allocation (pillar)",
  Strategy2a_ProportionalSubpillar = "Proportional allocation (sub-pillar)"
)
recode_strategy <- function(x) {
  out <- unname(strategy_labels[x])
  ifelse(is.na(out), x, out)
}

rand_flat_pillar    <- rand_index_flat$rand_index[rand_index_flat$comparison == "Full_vs_Strategy1_3perPillar"]
rand_flat_subpillar <- rand_index_flat$rand_index[rand_index_flat$comparison == "Full_vs_Strategy2_1perSubpillar"]
n_flat_subpillar    <- rand_index_flat$n_indicators_subset[rand_index_flat$comparison == "Full_vs_Strategy2_1perSubpillar"]

rand_proportional_pillar    <- rand_index_proportional$rand_index[rand_index_proportional$strategy == "Strategy1a_ProportionalPillar"]
rand_proportional_subpillar <- rand_index_proportional$rand_index[rand_index_proportional$strategy == "Strategy2a_ProportionalSubpillar"]

rand_index_flat_display <- rand_index_flat |>
  mutate(strategy = recode_strategy(strategy),
         comparison = paste("Full vs.", strategy))
rand_index_proportional_display <- rand_index_proportional |>
  mutate(strategy = recode_strategy(strategy),
         comparison = paste("Full vs.", strategy))

pca_components <- read.csv("outputs/pca_components.csv")
strategy2_var <- read.csv("outputs/variable_exhaustive_strategy2.csv")
strategy1_var <- read.csv("outputs/variable_exhaustive_strategy1.csv")
strategy1b_var <- read.csv("outputs/variable_exhaustive_strategy1b.csv")
strategy1a_var <- read.csv("outputs/variable_exhaustive_strategy1a.csv")
strategy2a_var <- read.csv("outputs/variable_exhaustive_strategy2a.csv")
results_by_dimension <- read.csv("outputs/results_by_dimension.csv")
r_m_curve <- read.csv("outputs/r_m_vs_pca_components.csv")
optimal_subset <- read.csv("outputs/optimal_21_variable_subset.csv")

imputation_report <- read.csv("data/imputation_report.csv")
original_stats <- read.csv("data/descriptive_stats_original.csv")
complete_stats <- read.csv("data/descriptive_stats_complete.csv")

fig2_k42 <- fig2_results[fig2_results$k == CONFIG$exhaustive$proportional_target, ]

# Step-wise local-search results: one row per (allocation, level, budget,
# metric) combination, and the long-format table of which indicator each
# combination selected from which group.
stepwise_results <- read.csv("outputs/stepwise_search_results.csv")
stepwise_vars <- read.csv("outputs/stepwise_selected_variables.csv")

# Full-dataset score a step-wise combination achieved, by combination code
# (e.g. "flat_pillar_21") and metric ("distance" or "r_m").
stepwise_score <- function(combination, metric_code) {
  stepwise_results$full_dataset_score[stepwise_results$combination == combination &
                                         stepwise_results$metric == metric_code]
}

# Indicator-overlap table for one perspective: for every group in `flat_df`
# (an exhaustive-search result table, e.g. strategy1_country), compare its
# selection to the exhaustive proportional result `proportional_df` and to
# the step-wise result named by `stepwise_combo` (run under the SAME
# allocation as `flat_df`, so the step-wise-vs-exhaustive comparison holds
# the per-group budget fixed). Reports, per group: the flat allocation's own
# local score (`score_col`, its only naturally group-level score - step-wise
# scores the whole candidate against the full dataset, not group-by-group -
# see Methodology), the indicators all three approaches agree on, the
# indicators shared by exactly two of the three (labeled with which pair),
# and the indicators unique to exactly one. Every indicator selected by any
# approach appears in exactly one of these six columns, so the row accounts
# for the full union of what was selected, not just the fully-unanimous or
# fully-unique picks.
build_overlap_table <- function(flat_df, proportional_df, stepwise_combo,
                                 metric_code, score_col, group_label_fn) {
  sw <- stepwise_vars[stepwise_vars$combination == stepwise_combo &
                         stepwise_vars$metric == metric_code, ]
  rows <- lapply(flat_df$group, function(g) {
    flat_set <- trimws(strsplit(flat_df$variables[flat_df$group == g], ",")[[1]])
    prop_set <- trimws(strsplit(proportional_df$variables[proportional_df$group == g], ",")[[1]])
    sw_set   <- sw$variable[sw$group == g]

    common    <- Reduce(intersect, list(flat_set, prop_set, sw_set))
    flat_only <- setdiff(flat_set, union(prop_set, sw_set))
    prop_only <- setdiff(prop_set, union(flat_set, sw_set))
    sw_only   <- setdiff(sw_set,   union(flat_set, prop_set))

    # Shared by exactly two of the three - would otherwise fall into neither
    # "common" (needs all three) nor any "-only" column (needs none of the
    # other two), leaving them silently unaccounted for.
    flat_prop <- setdiff(intersect(flat_set, prop_set), sw_set)
    flat_sw   <- setdiff(intersect(flat_set, sw_set), prop_set)
    prop_sw   <- setdiff(intersect(prop_set, sw_set), flat_set)
    shared_two <- c(
      if (length(flat_prop)) paste0(name_indicators_vec(flat_prop), " (Flat+Prop.)"),
      if (length(flat_sw))   paste0(name_indicators_vec(flat_sw), " (Flat+Step-wise)"),
      if (length(prop_sw))   paste0(name_indicators_vec(prop_sw), " (Prop.+Step-wise)")
    )

    data.frame(
      Group = group_label_fn(g),
      Score = round(flat_df[[score_col]][flat_df$group == g], 3),
      Common = paste(name_indicators_vec(common), collapse = "; "),
      SharedTwo = paste(shared_two, collapse = "; "),
      FlatOnly = paste(name_indicators_vec(flat_only), collapse = "; "),
      PropOnly = paste(name_indicators_vec(prop_only), collapse = "; "),
      StepwiseOnly = paste(name_indicators_vec(sw_only), collapse = "; "),
      stringsAsFactors = FALSE
    )
  })
  out <- do.call(rbind, rows)
  names(out) <- c("Group", "Within-Group Score (Flat)", "Common to All Three", "Shared by Two",
                   "Flat Only", "Proportional Only", "Step-wise Only")
  out
}
```

# Executive Summary

This report documents a method for determining how far the number of indicators in the Global Innovation Index (GII) 2025 can be reduced without significantly compromising the index's discriminatory power. The method was previously applied to the EU JRC's Sustainable and Inclusive Well-being Index [@benczur2024; @boskovic2024]; this report replicates it on GII's `r n_indicators_full` Level-1 indicators observed for `r n_countries` countries/economies. 

Two perspectives are evaluated: a **country perspective** (do countries end up the same distance apart using a subset as with the full indicator set?) and a **variable perspective** (does a subset preserve the PCA structure of the full set?). The two perspective rely on complementary metrics. The country perspective is based on a Spearman rank correlation between full- and subset-distance matrices; the variable perspective is based on $r_m^2$ - the percentage of the full PCA's variance structure recovered by the subset.

Both perspectives require the analyst to choose a subset of indicators. Subsets of 21 and 42 indicators are considered. As the GII has 21 sub-pillars, 21 is the minimum needed to cover every concept the index aims to measure.

Random selection provides the baseline: the best subset drawn from 1000 random selections at a given size, ignoring the conceptual structure entirely. Against it, four structured allocation rules are evaluated. Flat allocation assigns the same number of indicators to every group, either 3 per pillar or 1 per sub-pillar, both yielding 21 indicators. Proportional allocation instead sets each group's budget in proportion to its share of the index, on the premise that a group's indicator count reflects its importance or the difficulty of measuring it; this is applied across pillars and across sub-pillars at 42 indicators, and across pillars again at 21 so that the two rules can be compared at a common budget. Proportional allocation across sub-pillars at 21 indicators is not evaluated separately, because that apportionment reduces to exactly one indicator per sub-pillar and is therefore identical to the corresponding flat rule.

The the headline recommendation is that the "one indicator per sub-pillar" set of **`r nrow(strategy2_var)`- indicators comes close to reproducing the PCA structure** ($r_m$ = `r round(strategy2_var$r_m[1], 3)`, i.e. $r_m^2$ ≈ `r round(strategy2_var$r_m[1]^2 * 100, 1)`% of the full PCA's variance structure) while also closely reproducing the distance between countries (Rand Index `r round(rand_flat_subpillar, 3)` against the full-data country clustering). A larger `r sum(strategy1a_var$allocated)`-indicator subset, built by allocating indicators proportionally to each pillar's actual size rather than a flat count per pillar, reaches $r_m^2$ ≈ `r round(strategy1a_var$r_m[1]^2 * 100, 1)`% and a rank correlation of `r round(strategy1a$correlation[1], 3)` with the full country-distance matrix. Both are documented below alongside the flat and proportional strategies they were chosen from.

Relative to random subsets of matched size, exhaustive within-group search yields a modest advantage when indicators are allocated at pillar level and none when they are allocated at sub-pillar level. This pattern holds at both the `r nrow(strategy2_var)`- and `r CONFIG$exhaustive$proportional_target`-indicator budgets, and under both metrics.

# Introduction

When creating indicator dashboards and composite indices, indicator selection can be daunting. Stakeholders, statistics and conceptual relevance often diverge in recomendations or build up a large set of indicators that is impractical. A key statistical question is: how much of the information content ("variation") of the full indicator set is preserved in a candidate subset?

This analysis applies two methodological approaches to the GII 2025 data (`r n_indicators_full` Level-1 indicators, `r n_countries` countries):

1.  **Country Perspective**: Assessing how similar countries are using all indicators versus a subset, using distance matrices and clustering techniques
2.  **Variable Perspective**: Using Principal Component Analysis (PCA) to measure how much information a subset captures compared to the full set

## Finding a suitable reduced set

Composite indices such as the GII combine two different measurement logics. At the upper levels - pillars, sub-indices, the index itself - the structure is *formative*: each pillar captures a distinct dimension of innovation performance, and the index's meaning depends on all of them being represented; dropping a pillar entirely would change what the index measures, not just how precisely it measures it. Within a sub-pillar, by contrast, the indicators are closer to *reflective* measures of a single underlying concept - largely interchangeable manifestations of the same thing, where one or two well-chosen indicators can stand in for the group without changing what is being measured.

This distinction motivates the reduction strategy used throughout this analysis: every pillar and sub-pillar is retained in every candidate subset, preserving the formative structure and therefore the index's meaning, while indicator reduction happens only *within* each group, by selecting which reflective indicator(s) best represent it. We apply two strategies to decide *how many* reflective indicators to draw from each formative group. The first is to select one indicator per sub-pillar and the second is to select a number per pillar proportional to it's original size. The metrics for these two strategies ignore correlations between pillars or sub-pillars. However, these cannot be incorporated without also expanding the search to be exhaustive.

We have not implemented an exhaustive search for the best subset. Even restricted to searching *within* a single pillar, exhaustive search grows quickly. Selecting the best possible subset of size $k$ out of $p$ indicators requires evaluating $C(p,k) = \frac{p!}{k!(p-k)!}$ combinations. The `r pillar_names["IN.1"]` pillar (IN.1) has `r pillar_sizes["IN.1"]` indicators, so choosing 2 of them already means `r choose(pillar_sizes["IN.1"], 2)` possible combinations. Multiplied across all `r length(pillar_sizes)` pillars, the number of combinations for a subset is prohibitive. This analysis searches exhaustively within each pillar or sub-pillar rather than across the whole index at once. The table below shows how many combinations remain even after allocating a 20-indicator budget proportionally across pillars.

```{r combinations}
allocations |>
  transmute(
    Pillar = pillar_names[pillar],
    `Indicators in Pillar` = n_indicators,
    `Allocated (budget=20)` = allocation,
    Combinations = combinations
  ) |>
  knitr::kable(caption = "Unique combinations per pillar of GII when allocating a 20-indicator budget proportionally", digits = 2)
```

Even after this proportional allocation collapses the problem to one pillar at a time, there are still `r format(prod(allocations$combinations), big.mark = ",", scientific = FALSE)` combinations to explore in total. Underscoring why within-group exhaustive search is the only tractable approach.

# Methodology

## Statistical Methods Explanation

The variable perspective uses a subset-selection approach based on Principal Component Analysis (PCA). Rather than simply ranking individual variables by importance, this method evaluates entire subsets of variables according to how well they reproduce the PCA structure of the full dataset.

The criterion, $r_m$, was introduced by McCabe [-@mccabe1984] as one of several optimality criteria for selecting "principal variables". In contrast to principle components, principle variables are a subsets of the original variables that best capture variation in the original set. The method was developed in order to select indicators which could function as a principle component while remaining interpretable. Cadima and Jolliffe [-@cadima2001] formalize this same criterion as one of two indicators for how well a subset of variables approximates a full PCA, and show that $r_m^2$ is exactly the percentage of total variance in the original data accounted for by regressing the full data on the chosen subset. This is the interpretation of $r_2^m$ used throughout this report. Both papers show, that choosing variables by their PC loadings alone frequently produces subsets that are suboptimal compared to subsets chosen by directly maximizing $r_m$.

A key advantage of this method is its built-in protection against redundancy. For example, if multiple indicators are highly correlated, including one in the subset provides most of the information, and adding the others contributes relatively little additional ability to reconstruct the principal components.

The mathematical formula for the selection criterion is:

$$ r_m = \sqrt{ \frac{\sum_{i=1}^{p}\lambda_i r_{mi}^2} {\sum_{j=1}^{p}\lambda_j} }$$

where: - $\lambda_i$ is the eigenvalue of principal component $i$ ( $PC_{i}$ ) from the full dataset, representing the variance explained by that component. The subscript $m$ indexes a candidate subset - $r_{mi}^2$ is the coefficient of determination ($R^2$) between principal component $i$ and the candidate subset of variables. Since $PC_{i}$ is a linear combination of the original variables, the $R^2$ of the subset measures how well the principal component can be reconstructed from the subset.

Conceptually, this formula computes a weighted average of how well the subset predicts each principal component, with weights given by the variance explained by each component, then normalizes by the total variance. The resulting $r_m^2$ value can be interpreted as the percentage of total variation explained by the subset, providing an intuitive measure of information retention.

This method assumes that "information" is defined by the variance structure captured by PCA, which is appropriate for statistical data reduction. However, it may downweight variables with co-variance but high policy relevance, which is an important consideration for substantive index development (c.f.[@hleg2026]).

### The Country Perspective's Distance-Based Criterion

The country perspective asks assess whether a subset preserves how similar countries are judged to be relative to one another. For each candidate subset, the pairwise Euclidean distance between every pair of countries is computed using only that subset's indicators, and the ranking of those pairwise distances is compared to the ranking obtained from the full indicator set, via Spearman rank correlation. A correlation near 1 means the subset reproduces almost the same country-similarity ordering as the full index.

### Step-wise Local Search

Exhaustive search identifies the single best-scoring subset within a group, but it is only tractable because it is applied separately to each pillar or sub-pillar; the criterion it maximizes for a given group is therefore local to that group's own indicators, rather than the full index. A step-wise (greedy) local-search procedure offers an alternative that scales to a group or a budget too large to search exhaustively, and that optimizes a candidate subset's score against the FULL indicator set directly, rather than against any single group in isolation. It applies equally to either selection criterion - the country perspective's rank correlation or the variable perspective's $r_m$ - and to any allocation of indicators to groups.

The procedure proceeds as follows:

1.  **Initialization.** For every group, the number of indicators specified by that group's allocation is drawn at random, without replacement, from the indicators belonging to that group. Together these draws form the initial candidate subset.
2.  **Scoring.** The candidate subset is scored against the full indicator set - by the rank correlation of distance matrices, or by $r_m$ against the full-dataset PCA - not against any single group's own indicators.
3.  **Local search.** Each indicator currently in the subset is considered for replacement, in turn, by every other indicator belonging to the same group. A replacement is accepted whenever it raises the subset's full-dataset score, and search continues from the improved subset; a replacement that would introduce an indicator already present elsewhere in the subset is never considered, since a duplicated indicator would be double-counted in the underlying distance or PCA calculation.
4.  **Convergence.** Step 3 is repeated, as a full pass over every group and every indicator in it, until a complete pass produces no further improvement, or a fixed maximum number of passes is reached.

Because the search starts from a random draw and only ever accepts a strictly improving substitution, it converges to a local optimum with respect to one-indicator-at-a-time swaps, which need not be the global optimum an exhaustive search would find were one computationally feasible. The procedure is therefore run from several independent random starting points, and the best-scoring outcome across those runs is retained.

## Proportional Allocation Methodology

Both perspectives select candidate subsets by exhaustive search *within* each pillar or sub-pillar, then assemble the per-group picks into a full subset - consistent with the formative/reflective distinction. Every pillar and sub-pillar is represented in every candidate subset, and only the reflective indicators within each group are subject to selection. This requires deciding how many indicators to take from each group. Two allocation rules are used throughout this report:

- **Flat allocation of 21 indicators**: every pillar contributes (`r CONFIG$exhaustive$strategy1_k_per_pillar`) indicators, or every sub-pillar contributes `r CONFIG$exhaustive$strategy2_k_per_subpillar` indicators. A simple strategy to implement, however, it ignores that GII's pillars differ in the number of indicators, from `r min(pillar_sizes)` indicators (`r pillar_names[names(which.min(pillar_sizes))]`) to `r max(pillar_sizes)` (`r pillar_names[names(which.max(pillar_sizes))]`), and its sub-pillars range from `r min(subpillar_sizes)` to `r max(subpillar_sizes)` indicators. A flat allocation over-represents small groups and under-represents large ones relative to their actual share of the index. This may limit the subset's utility statistically and leave out conceptual nuances of the GII measurement framework.
- **Proportional allocation of 21 or 42 indicators**. Each group's budget is set proportional to its share of all indicators, using a **largest-remainder (Hamilton) apportionment**. Every group first receives *n* indicators rounded down to the lowest integer. Any indicators left over are handed out one at a time to the groups with the largest fractional remainder. Unlike flat allocation, a group with a very small share of the index can receive **zero** indicators under this rule. The total budget used is **`r CONFIG$exhaustive$proportional_target`** indicators (double the flat allocations' `r CONFIG$exhaustive$strategy1_k_per_pillar * length(pillar_sizes)`). Proportional allocation is applied across pillars and across sub-pillars at this budget, and across pillars again at **`r CONFIG$exhaustive$proportional_target_matched`** indicators, matching the flat allocations so that the effect of the allocation rule can be separated from the effect of the budget. No matching sub-pillar case exists: apportioning `r CONFIG$exhaustive$proportional_target_matched` indicators across `r length(subpillar_sizes)` sub-pillars returns exactly one indicator each, which is the flat rule. This is also why the larger budget is set above the number of sub-pillars.

Each perspective searches on its own criterion, applied *locally* within a group. The variable perspective maximizes $r_m$ against that group's own sub-PCA rather than the full `r n_indicators_full`-indicator PCA. The country perspective maximizes the Spearman rank correlation between that group's own distance matrix and the candidate subset's. The allocation rule and the selection criterion are independent choices, so all four allocation rules above are searched under both criteria. This local search is a tractable stand-in for an otherwise infeasible full search. The assembled subset is then re-scored against the full dataset to obtain the metrics reported in this document.

## Data Description

The Global Innovation Index (GII) 2025 dataset [@wipo2025] comprises `r n_indicators_full` Level-1 indicators observed for `r n_countries` countries/economies, organized in a 5-level hierarchy: `r n_indicators_full` indicators -\> `r length(subpillar_sizes)` sub-pillars -\> `r length(pillar_sizes)` pillars -\> `r n_subindices` sub-indices (Inputs, Outputs) -\> `r n_index` overall index. Indicators with negative polarity (e.g. pupil-teacher ratio, applied tariff rate) are reversed before standardization, so that higher standardized values consistently indicate better innovation performance across all indicators. Missing values are imputed via 5-nearest-neighbors using the VIM [@Kowarik2016] package prior to z-scoring; see the Appendix for imputation statistics.

# Key Findings

## Country Perspective

### Distance Matrix Analysis of Random Subsets

We selected random subset sizes ranging from 10% to 90% of all `r n_indicators_full` indicators, while also selecting specifically `r CONFIG$exhaustive$proportional_target`, matching the proportional strategies' budget below. Each k subset was drawn 1000 times and compared to the full-dataset Euclidean distance matrix via Spearman rank correlation:

```{r}
fig2_results |>
  rename(`Subset Size (k)` = k, `Mean Correlation` = mean_correlation,
         `Median Correlation` = median_correlation, `Min Correlation` = min_correlation,
         `Max Correlation` = max_correlation) |>
  knitr::kable(caption = "Rank correlation of random subsets vs. the full distance matrix, by subset size k", digits = 3)
```

Correlation rises smoothly and near-monotonically with subset size, from a mean of `r round(fig2_results$mean_correlation[1], 3)` at k = `r fig2_results$k[1]` to `r round(fig2_results$mean_correlation[nrow(fig2_results)], 3)` at k = `r fig2_results$k[nrow(fig2_results)]`. This random-subset curve is the baseline to the local exhaustive-search strategies.

### Flat Allocation

**Finding:** Flat allocation of `r CONFIG$exhaustive$strategy1_k_per_pillar` indicators per pillar reaches a rank correlation of `r round(strategy1_country$correlation[1], 3)` with the full distance matrix; flat allocation of `r CONFIG$exhaustive$strategy2_k_per_subpillar` indicator per sub-pillar reaches `r round(strategy2_country$correlation[1], 3)`. The comparable random baseline (k = 23) averages `r round(fig2_results$mean_correlation[fig2_results$k == 23], 3)`, with a maximum of `r round(fig2_results$max_correlation[fig2_results$k == 23], 3)` across 1000 draws.

The two allocation levels therefore diverge. Pillar-level allocation exceeds the random average by `r round(strategy1_country$correlation[1] - fig2_results$mean_correlation[fig2_results$k == 23], 3)`, while sub-pillar-level allocation falls `r round(fig2_results$mean_correlation[fig2_results$k == 23] - strategy2_country$correlation[1], 3)` below it. Neither reaches the best random subset of the same size, which is expected: the search is confined within groups, so it cannot exploit the cross-group combinations available to an unconstrained draw. The constraint is accepted deliberately, since it is what guarantees every pillar and sub-pillar remains represented and the index retains its formative meaning.

### Clustering Results: SIMPROF + Rand Index

Against the full-data country clustering, flat allocation by pillar recovers a Rand Index of `r round(rand_flat_pillar, 3)` and flat allocation by sub-pillar `r round(rand_flat_subpillar, 3)`. Both subsets therefore reproduce the full clustering to within roughly `r round((1 - min(rand_flat_pillar, rand_flat_subpillar)) * 100)`% of country pairs, and the two allocation levels are indistinguishable on this measure despite their `r round(strategy1_country$correlation[1] - strategy2_country$correlation[1], 3)` gap in rank correlation.

```{r}
rand_index_flat_display |>
  rename(Comparison = comparison, `Rand Index` = rand_index, `Full Indicator Count` = n_indicators_full,
         `Subset Indicator Count` = n_indicators_subset, `Allocation Rule` = strategy) |>
  knitr::kable(caption = "Rand Index: full-data clustering vs. flat-allocation clustering (by pillar and by sub-pillar)", digits = 3)
```

### Proportional Allocation

We proportionally allocate `r CONFIG$exhaustive$proportional_target` indicators, using the largest-remainder method described in Methodology above.

**Finding:** At the `r CONFIG$exhaustive$proportional_target`-indicator budget, proportional allocation across pillars achieves a rank correlation of `r round(strategy1a$correlation[1], 3)`, against `r round(strategy2a$correlation[1], 3)` across sub-pillars. The random baseline at this size averages `r round(fig2_k42$mean_correlation[1], 3)`.

The advantage of pillar-level over sub-pillar-level allocation seen at 21 indicators persists at 42, and in the same direction: pillar-level allocation again exceeds the random average while sub-pillar-level allocation falls below it. Doubling the budget raises every figure, but does not alter the ordering between the two allocation levels.

Applying the same proportional rule across pillars at the flat `r CONFIG$exhaustive$proportional_target_matched`-indicator budget returns `r round(strategy1b$correlation[1], 3)`, against `r round(strategy1_country$correlation[1], 3)` for flat allocation at that budget. Separating the two effects this way shows the gain to be a function of the budget rather than of the rule.

```{r}
rand_index_proportional_display |>
  rename(Comparison = comparison, `Rand Index` = rand_index, `Full Indicator Count` = n_indicators_full,
         `Subset Indicator Count` = n_indicators_subset, `Allocation Rule` = strategy) |>
  knitr::kable(caption = "Rand Index: full clustering vs. proportional-allocation subset clustering (by pillar and by sub-pillar)", digits = 3)
```

**Finding:** Against the full-data country clustering, the `r CONFIG$exhaustive$proportional_target`-indicator proportional subset recovers a Rand Index of `r round(rand_proportional_pillar, 3)` when allocated across pillars and `r round(rand_proportional_subpillar, 3)` when allocated across sub-pillars - against `r round(rand_flat_pillar, 3)` and `r round(rand_flat_subpillar, 3)` for the two `r nrow(strategy2_country)`-indicator flat subsets.

All four values fall within `r round(max(c(rand_flat_pillar, rand_flat_subpillar, rand_proportional_pillar, rand_proportional_subpillar)) - min(c(rand_flat_pillar, rand_flat_subpillar, rand_proportional_pillar, rand_proportional_subpillar)), 3)` of one another, even though the subsets differ in size by a factor of two and span `r round(strategy1a$correlation[1] - strategy2_country$correlation[1], 3)` in rank correlation. Recovery of the country clustering is therefore insensitive to both subset size and allocation level over this range: once every pillar and sub-pillar is represented, additional indicators sharpen the distance structure without redrawing the cluster boundaries.

### Step-wise Local Search

The step-wise local-search procedure (Methodology) was run under the rank-correlation criterion for every allocation rule (flat, proportional), grouping level (pillar, sub-pillar) and budget (21, 42) considered above, with `r unique(stepwise_results$n_restarts)` independent random restarts per combination and the best-scoring restart retained.

```{r}
stepwise_results |>
  filter(metric == "distance") |>
  transmute(
    Level = ifelse(level == "pillar", "Pillar", "Sub-pillar"),
    Budget = budget,
    `Allocation Rule` = ifelse(allocation_type == "flat", "Flat", "Proportional"),
    `Rank Correlation` = full_dataset_score
  ) |>
  arrange(Level, Budget, `Allocation Rule`) |>
  knitr::kable(caption = "Step-wise search: rank correlation with the full distance matrix, by allocation", digits = 3, row.names = FALSE)
```

**Finding:** At the `r nrow(strategy2_var)`-indicator budget, step-wise search reaches a rank correlation of `r round(stepwise_score("flat_pillar_21", "distance"), 3)` under the flat pillar allocation and `r round(stepwise_score("prop_pillar_21", "distance"), 3)` under the proportional pillar allocation - both above the exhaustive-search results for the same allocations (`r round(strategy1_country$correlation[1], 3)` and `r round(strategy1b$correlation[1], 3)` respectively) and above the random baseline at matched size (`r round(fig2_results$mean_correlation[fig2_results$k == 23], 3)`). Under the flat sub-pillar allocation, step-wise search reaches `r round(stepwise_score("flat_subpillar_21", "distance"), 3)`, again exceeding both the exhaustive result (`r round(strategy2_country$correlation[1], 3)`) and the random baseline - reversing the earlier finding that sub-pillar-level allocation offers no advantage over chance. The same pattern - step-wise above both its exhaustive-search and random-baseline counterparts - holds at the `r CONFIG$exhaustive$proportional_target`-indicator budget across every allocation tested (Table above).

This gap is a direct consequence of what each method optimizes. Exhaustive search maximizes each group's *local* score, computed only against that group's own indicators; the full-dataset score reported for it is a byproduct, not the search objective (see Limitations). Step-wise search instead scores every candidate against the full dataset at every step, so it can find combinations that a group-by-group local search cannot see - at the cost of no longer guaranteeing that the result is the best subset achievable for that allocation, only a local optimum with respect to one-indicator-at-a-time swaps from a random start.

The table below breaks down, pillar by pillar, where the three approaches agree and diverge, for the `r CONFIG$exhaustive$strategy1_k_per_pillar`-indicators-per-pillar budget flat allocation uses: flat allocation's own exhaustive result, proportional allocation at the same matched budget, and step-wise search run under the flat allocation (so the step-wise-vs-exhaustive columns hold the per-pillar budget fixed). Every selected indicator appears in exactly one column: "Common to All Three" for indicators every approach picked, "Shared by Two" (labeled with which pair) for indicators exactly two agreed on, and one of the three "-only" columns for indicators picked by exactly one. A parallel breakdown at sub-pillar level is not shown: at this budget, proportional allocation across sub-pillars is mathematically identical to flat allocation (Methodology), so the three-way comparison would collapse to flat vs. step-wise alone, and the sub-pillar-level scores already appear in the table above.

\blandscape

```{r}
build_overlap_table(strategy1_country, strategy1b, "flat_pillar_21", "distance",
                     "best_spearman", function(x) pillar_names[x]) |>
  knitr::kable(caption = "Country perspective: indicator overlap between flat allocation, proportional allocation, and step-wise search, by pillar (21-indicator budget)", row.names = FALSE)
```

\elandscape

## Variable Perspective

### PCA Analysis

The table below shows the number of principal components needed to capture specified levels of variance in the full `r n_indicators_full`-indicator dataset:

```{r}
pca_components |>
  rename(`Variance Level` = Variance_Level, `Components Needed` = Components_Needed) |>
  knitr::kable(caption = "Components needed for target variance levels (full PCA)")
```

**Finding:** The first `r pca_components[1,2]` of `r n_indicators_full` components already capture 80% of total variance, confirming substantial redundancy in the full indicator set - a necessary condition for indicator reduction to be viable at all. Note this is a property of the full PCA (unconstrained linear combinations of all `r n_indicators_full` indicators), not directly comparable to how many *raw indicators* a subset needs.

### $r_m$ by Group and by Strategy

The exhaustive-search sections below each report a *global* $r_m$ - the assembled subset's fit to the full-dataset PCA - built up from a *local* $r_m$ computed separately for every pillar or sub-pillar. This section collects both views into two reference tables before the individual strategies are discussed one at a time.

The table below reports flat allocation's local $r_m$ - the score its exhaustive search actually maximizes - for every pillar and every sub-pillar (the same per-pillar values that also appear as the "Within-Group Score (Flat)" column of the indicator-overlap table under Step-wise Local Search, below):

```{r}
rm_by_group <- bind_rows(
  strategy1_var |> transmute(Level = "Pillar", Group = pillar_names[group], `$r_m$` = best_rm),
  strategy2_var |> transmute(Level = "Sub-pillar", Group = subpillar_names[group], `$r_m$` = best_rm)
)
knitr::kable(rm_by_group, digits = 3, caption = "Within-group $r_m$ (flat allocation's exhaustive search), by pillar and sub-pillar", row.names = FALSE)
```

### Comparison of flat and proportional strategies in gobal $r_m$ 

The table below instead combines the *global* $r_m$ reported by each of the flat and proportional strategies discussed in the following sections into a single comparison (the step-wise alternative to these is introduced later, and appears in the Cross-Allocation Comparison table with the rest):

```{r}
global_rm_summary <- data.frame(
  Strategy = c("Flat allocation (sub-pillar, headline)", "Flat allocation (pillar)",
               "Proportional allocation (pillar)", "Proportional allocation (sub-pillar)"),
  Indicators = c(nrow(strategy2_var), nrow(strategy1_var) * CONFIG$exhaustive$strategy1_k_per_pillar,
                 sum(strategy1a_var$allocated), sum(strategy2a_var$allocated)),
  `$r_m$` = c(strategy2_var$r_m[1], strategy1_var$r_m[1], strategy1a_var$r_m[1], strategy2a_var$r_m[1]),
  check.names = FALSE
)
global_rm_summary$`$r_m^2$ (%)` <- round(global_rm_summary$`$r_m$`^2 * 100, 1)
knitr::kable(global_rm_summary, digits = 3, caption = "Global $r_m$ (assembled subset vs. full-dataset PCA), by strategy", row.names = FALSE)
```

**Finding:** At `r sum(strategy1a_var$allocated)` indicators, proportional allocation by pillar reaches $r_m$ = `r round(strategy1a_var$r_m[1], 3)` ($r_m^2$ ≈ `r round(strategy1a_var$r_m[1]^2 * 100, 1)`%) and proportional allocation by sub-pillar reaches $r_m$ = `r round(strategy2a_var$r_m[1], 3)` ($r_m^2$ ≈ `r round(strategy2a_var$r_m[1]^2 * 100, 1)`%) - both a large step up from the `r nrow(strategy2_var)`-indicator flat allocations, as expected with double the budget.

### Step-wise Local Search

The same step-wise local-search procedure (Methodology) was run under the $r_m$ criterion, for every allocation rule, grouping level and budget considered above, with `r unique(stepwise_results$n_restarts)` independent random restarts per combination and the best-scoring restart retained.

```{r}
stepwise_results |>
  filter(metric == "r_m") |>
  transmute(
    Level = ifelse(level == "pillar", "Pillar", "Sub-pillar"),
    Budget = budget,
    `Allocation Rule` = ifelse(allocation_type == "flat", "Flat", "Proportional"),
    `$r_m$` = full_dataset_score
  ) |>
  arrange(Level, Budget, `Allocation Rule`) |>
  knitr::kable(caption = "Step-wise search: $r_m$ against the full-dataset PCA, by allocation", digits = 3, row.names = FALSE)
```

**Finding:** At `r nrow(strategy2_var)` indicators, step-wise search reaches $r_m$ = `r round(stepwise_score("flat_pillar_21", "r_m"), 3)` under the flat pillar allocation and `r round(stepwise_score("prop_pillar_21", "r_m"), 3)` under the proportional pillar allocation - both above the corresponding exhaustive-search results (`r round(strategy1_var$r_m[1], 3)` and `r round(strategy1b_var$r_m[1], 3)`) and above the random-subset baseline at matched size (`r round(r_m_curve$mean_r_m[r_m_curve$k == 23], 3)`). Under the flat sub-pillar allocation, step-wise search reaches $r_m$ = `r round(stepwise_score("flat_subpillar_21", "r_m"), 3)`, again exceeding both the exhaustive result (`r round(strategy2_var$r_m[1], 3)`) and the random baseline - reversing the earlier finding that, at this budget, the constrained search offers no advantage over chance on the PCA criterion. The same pattern holds at `r CONFIG$exhaustive$proportional_target` indicators across every allocation tested (Table above), for the same reason given in the country perspective: step-wise search directly optimizes the full-dataset $r_m$ at every step, where exhaustive search optimizes each group's local $r_m$ and is only re-scored against the full PCA afterwards (Limitations).

As in the country perspective, the table below compares flat allocation's exhaustive result, proportional allocation at the matched budget, and step-wise search run under the flat allocation, pillar by pillar, with every selected indicator accounted for in exactly one of the "Common to All Three", "Shared by Two", or single-approach "-only" columns.

\blandscape

```{r}
build_overlap_table(strategy1_var, strategy1b_var, "flat_pillar_21", "r_m",
                     "best_rm", function(x) pillar_names[x]) |>
  knitr::kable(caption = "Variable perspective: indicator overlap between flat allocation, proportional allocation, and step-wise search, by pillar (21-indicator budget)", row.names = FALSE)
```

\elandscape

### Flat-Allocation Sweep: 1-4 Indicators per Pillar

To see how $r_m$ grows as the flat per-pillar allocation increases, the flat allocation by pillar search was repeated for 1, 2, 3, and 4 indicators per pillar:

```{r}
results_by_dimension |>
  distinct(n_indicators, total_rm) |>
  mutate(n_variables = n_indicators * length(pillar_sizes)) |>
  rename(`Indicators per Pillar` = n_indicators, `$r_m$` = total_rm, `Total Indicators` = n_variables) |>
  knitr::kable(caption = "$r_m$ as the flat per-pillar allocation increases", digits = 3)

sweep_summary <- results_by_dimension |>
  distinct(n_indicators, total_rm) |>
  mutate(n_variables = n_indicators * length(pillar_sizes))
```

**Finding:** $r_m$ rises from `r round(sweep_summary$total_rm[1], 3)` at `r sweep_summary$n_variables[1]` variables (1/pillar) to `r round(sweep_summary$total_rm[4], 3)` at `r sweep_summary$n_variables[4]` variables (4/pillar) - a steep gain per additional variable early on, consistent with the PCA finding above that a small number of dimensions already captures most of the variance structure.

### Random-Subset $r_m$ Curve vs. #PCs Needed

For the same subset sizes used in the country perspective's Fig 2, $r_m$ was computed for `r CONFIG$sampling$n_random_subsets` random subsets per size. $r_m$ is far more expensive per evaluation than the rank-correlation metric. It requires regressing all `r n_indicators_full` principal components on the candidate subset, so fewer random draws were used per size here than Table 2's 1,000.

```{r}
r_m_curve |>
  rename(`Subset Size (k)` = k, `$r_m$` = mean_r_m, `Subset Variance Captured` = subset_variance_captured,
         `PCA Variance (k components)` = pca_variance_at_k_components) |>
  knitr::kable(caption = "Random-subset $r_m$ by size, vs. cumulative variance explained by the first k principal components", digits = 3)
```

**Finding:** At k = `r nrow(strategy2_var)` (interpolating between k = 15 and k = 23 above), a random subset averages roughly the same $r_m$ as the exhaustive flat-allocation results report (`r round(strategy1_var$r_m[1], 3)` by pillar / `r round(strategy2_var$r_m[1], 3)` by sub-pillar) - so at this budget the constrained search offers no advantage over chance on the PCA criterion, in contrast to the country perspective, where pillar-level allocation exceeded its random baseline. The PCA-variance column is consistently higher than a same-sized *indicator* subset's $r_m^2$ at every k. Since the first k principal components are unconstrained linear combinations of all `r n_indicators_full` indicators, a subset is restricted to k raw indicators. The resulting gap in explained variance is the cost of requiring the final subset to be interpretable indicators rather than statistical composites.

### Headline Recommended Subset

The `r nrow(strategy2_var)`-indicator subset from flat allocation by sub-pillar above, with full indicator names:

```{r}
optimal_subset |>
  transmute(
    `Sub-Pillar` = subpillar_names[subpillar],
    Indicator = indicator_names[variable],
    `$r_m$ of Best Subset` = subpillar_r_m
  ) |>
  knitr::kable(caption = "Headline subset: one reflective indicator per sub-pillar", digits = 3)
```

**Finding:** This subset captures $r_m^2$ ≈ `r round(optimal_subset$overall_r_m[1]^2 * 100, 1)`% of the full dataset's PCA variance structure and recovers a Rand Index of `r round(rand_flat_subpillar, 3)` against the full-data country clustering, while cutting the indicator count from `r n_indicators_full` to `r nrow(optimal_subset)` (a `r round((1 - nrow(optimal_subset) / n_indicators_full) * 100)`% reduction).

# Cross-Allocation Comparison

The tables below collapse every allocation rule, grouping level and budget considered above - under both exhaustive search and step-wise search - into a single comparison per perspective.

```{r}
exhaustive_country <- data.frame(
  Level = c("Pillar", "Sub-pillar", "Pillar", "Pillar", "Sub-pillar"),
  `Allocation Rule` = c("Flat", "Flat", "Proportional", "Proportional", "Proportional"),
  Indicators = c(nrow(strategy1_country) * CONFIG$exhaustive$strategy1_k_per_pillar, nrow(strategy2_country),
                 sum(strategy1b$allocated), sum(strategy1a$allocated), sum(strategy2a$allocated)),
  `Rank Correlation` = c(strategy1_country$correlation[1], strategy2_country$correlation[1],
                         strategy1b$correlation[1], strategy1a$correlation[1], strategy2a$correlation[1]),
  `Search Method` = "Exhaustive search",
  check.names = FALSE
)
stepwise_country <- stepwise_results |>
  filter(metric == "distance") |>
  transmute(
    Level = ifelse(level == "pillar", "Pillar", "Sub-pillar"),
    `Allocation Rule` = ifelse(allocation_type == "flat", "Flat", "Proportional"),
    Indicators = budget,
    `Rank Correlation` = full_dataset_score,
    `Search Method` = "Step-wise search"
  )
country_comparison <- rbind(exhaustive_country, stepwise_country)
country_comparison <- country_comparison[order(country_comparison$Level, country_comparison$Indicators,
                                                 country_comparison$`Allocation Rule`, country_comparison$`Search Method`), ]
knitr::kable(country_comparison, caption = "Country perspective: allocation rules compared, by level, budget and search method",
             digits = 3, row.names = FALSE)
```

```{r}
exhaustive_var <- data.frame(
  Level = c("Pillar", "Sub-pillar", "Pillar", "Pillar", "Sub-pillar"),
  `Allocation Rule` = c("Flat", "Flat", "Proportional", "Proportional", "Proportional"),
  Indicators = c(nrow(strategy1_var) * CONFIG$exhaustive$strategy1_k_per_pillar, nrow(strategy2_var),
                 sum(strategy1b_var$allocated), sum(strategy1a_var$allocated), sum(strategy2a_var$allocated)),
  r_m = c(strategy1_var$r_m[1], strategy2_var$r_m[1],
          strategy1b_var$r_m[1], strategy1a_var$r_m[1], strategy2a_var$r_m[1]),
  `Search Method` = "Exhaustive search",
  check.names = FALSE
)
stepwise_var <- stepwise_results |>
  filter(metric == "r_m") |>
  transmute(
    Level = ifelse(level == "pillar", "Pillar", "Sub-pillar"),
    `Allocation Rule` = ifelse(allocation_type == "flat", "Flat", "Proportional"),
    Indicators = budget,
    r_m = full_dataset_score,
    `Search Method` = "Step-wise search"
  )
variable_comparison <- rbind(exhaustive_var, stepwise_var)
variable_comparison <- variable_comparison[order(variable_comparison$Level, variable_comparison$Indicators,
                                                   variable_comparison$`Allocation Rule`, variable_comparison$`Search Method`), ]
variable_comparison$r_m_squared_pct <- round(variable_comparison$r_m^2 * 100, 1)
names(variable_comparison) <- c("Level", "Allocation Rule", "Indicators", "$r_m$", "Search Method", "$r_m^2$ (%)")
variable_comparison <- variable_comparison[, c("Level", "Allocation Rule", "Indicators", "$r_m$", "$r_m^2$ (%)", "Search Method")]
knitr::kable(variable_comparison, caption = "Variable perspective: allocation rules compared, by level, budget and search method", digits = 3, row.names = FALSE)
```

**Finding:** Step-wise search scores at or above its exhaustive-search counterpart in every one of the five allocation/budget combinations both methods were run on, under both metrics (ten comparisons in total) - not a coincidence, but a consequence of what each method optimizes: exhaustive search maximizes a *local*, group-level criterion and is merely re-scored against the full dataset afterwards, while step-wise search optimizes the full-dataset criterion directly at every step (Methodology; discussed per-perspective above). Exhaustive search remains the only one of the two with a guarantee of finding the best possible subset for the criterion it is given; step-wise search offers no such guarantee, only a local optimum from a random start, mitigated here by keeping the best of several restarts.

**Finding:** Proportional allocation at `r CONFIG$exhaustive$proportional_target` indicators outperforms flat allocation at `r nrow(strategy2_var)` indicators on both metrics. Compared at a common budget, however, the two rules are nearly indistinguishable: at `r nrow(strategy2_var)` indicators across pillars, proportional allocation reaches a rank correlation of `r round(strategy1b$correlation[1], 3)` against flat allocation's `r round(strategy1_country$correlation[1], 3)`, and $r_m$ = `r round(strategy1b_var$r_m[1], 3)` against `r round(strategy1_var$r_m[1], 3)`.

Budget therefore accounts for almost all of the apparent advantage. Holding the rule fixed and doubling the budget from `r nrow(strategy2_var)` to `r CONFIG$exhaustive$proportional_target` indicators raises the rank correlation by `r round(strategy1a$correlation[1] - strategy1b$correlation[1], 3)` and $r_m$ by `r round(strategy1a_var$r_m[1] - strategy1b_var$r_m[1], 3)`; holding the budget fixed and switching from flat to proportional changes them by `r round(strategy1b$correlation[1] - strategy1_country$correlation[1], 3)` and `r round(strategy1b_var$r_m[1] - strategy1_var$r_m[1], 3)`. The budget effect is roughly two orders of magnitude larger on the distance metric and one on the PCA metric.

The case for proportional allocation is therefore representational rather than statistical: no group is systematically over- or under-weighted relative to its size in the index, at no measurable cost to either criterion. Whether indicator count is an appropriate basis for that weighting remains a judgement for the analyst, and the near-equivalence observed here need not hold for an index whose sub-pillars are more strongly correlated within pillars.

# Conclusions

1.  **Meaningful reduction is possible, with a clear methodological trade-off.** The headline `r nrow(strategy2_var)`-indicator subset (one reflective indicator per sub-pillar) captures $r_m^2$ ≈ `r round(optimal_subset$overall_r_m[1]^2 * 100, 1)`% of the full PCA's variance structure and a Rand Index of `r round(rand_flat_subpillar, 3)` against the full clustering, at a `r round((1 - nrow(optimal_subset) / n_indicators_full) * 100)`% reduction in indicator count.

2.  **The gain over random selection depends on the level at which indicators are allocated, not on the budget.** Allocating at pillar level exceeds the average random subset of matched size at both budgets (`r round(strategy1_country$correlation[1], 3)` against `r round(fig2_results$mean_correlation[fig2_results$k == 23], 3)` at `r nrow(strategy2_var)` indicators; `r round(strategy1a$correlation[1], 3)` against `r round(fig2_k42$mean_correlation[1], 3)` at `r CONFIG$exhaustive$proportional_target`). Allocating at sub-pillar level falls below it at both. No constrained subset reaches the best random subset of its size, which is the cost of requiring every group to be represented.

3.  **The allocation rule matters far less than the budget.** Compared at a common `r nrow(strategy2_var)`-indicator budget, flat and proportional allocation across pillars differ by `r round(strategy1b$correlation[1] - strategy1_country$correlation[1], 3)` in rank correlation and `r round(strategy1b_var$r_m[1] - strategy1_var$r_m[1], 3)` in $r_m$, while doubling the budget under a fixed rule gains `r round(strategy1a$correlation[1] - strategy1b$correlation[1], 3)` and `r round(strategy1a_var$r_m[1] - strategy1b_var$r_m[1], 3)` respectively. Proportional allocation's benefit is representational rather than statistical.

4.  **Recovery of the country clustering saturates well below the full indicator set.** The four candidate subsets (flat and proportional, by pillar and by sub-pillar) return Rand Indices within `r round(max(c(rand_flat_pillar, rand_flat_subpillar, rand_proportional_pillar, rand_proportional_subpillar)) - min(c(rand_flat_pillar, rand_flat_subpillar, rand_proportional_pillar, rand_proportional_subpillar)), 3)` of one another despite differing twofold in size and by `r round(strategy1a$correlation[1] - strategy2_country$correlation[1], 3)` in rank correlation. Beyond covering every pillar and sub-pillar, further indicators refine the distance structure without changing which countries group together.

5.  **The** $r_m$ criterion treats "information" as PCA variance, which may downweight indicators with low covariance but high substantive/policy relevance; the headline subset should be reviewed against domain expertise, not adopted on statistical grounds alone.

# Limitations

1.  **Single-year analysis.** This analysis covers GII 2025 only; temporal stability of the recommended subset across years has not been assessed.
2.  **PCA-defined "information."** The $r_m$ criterion equates information content with PCA variance structure; a substantively important but low covariance indicator could be excluded even if policy-relevant.
3.  **Local search proxy.** Exhaustive search within each pillar/sub-pillar uses a local $r_m$ computed on that group's own sub-PCA (not the full `r n_indicators_full`-indicator PCA) as a tractable stand-in for an otherwise infeasible full search; the assembled subset is re-scored against the true full-dataset metric for reporting, but the *search* itself never directly optimizes that metric. Step-wise local search (Methodology) avoids this specific limitation, since it optimizes the full-dataset metric directly - at the cost of no longer guaranteeing the result is the best subset achievable for that allocation, only a local optimum from a random start.
4.  **Linear relationships assumed.** Both the rank-correlation and $r_m$ criteria are built on linear/monotonic structure (PCA, Euclidean distance); non-linear relationships between indicators are not captured.
5.  **Domain-specific validity not tested.** Whether the recommended subset remains meaningful to GII's intended policy audience has not been evaluated here and would require domain-expert review.

# Data Availability

All data and code are available in this repository:

- Raw data: `gii-data/idata.xlsx`, `gii-data/imeta.xlsx`
- Processed data: `data/gii_data_standardized.csv`, `data/gii_data_complete_cases.csv`
- Analysis scripts: `scripts/00_setup.R` through `scripts/04_variable_perspective.R`, plus `scripts/04_step-wise.R` (step-wise local search), shared functions in `scripts/R/`
- Results: `outputs/` (CSVs for tabular results, `.rds` for clustering objects; `stepwise_search_results.csv` and `stepwise_selected_variables.csv` for the step-wise search)

# References

::: {#refs}
:::

# Appendix

# Data Imputation Documentation

Missing data is an inherent challenge in composite index construction. This analysis used k-Nearest Neighbors (kNN) imputation to address missing values before conducting the main analyses.

## Imputation Process

- **Method**: k-Nearest Neighbors (kNN) imputation
- **Parameters**: k = 5 neighbors (`VIM::kNN`, other parameters left at package defaults)
- **Implementation**: R `VIM` package
- **Indicators Used**: Level 1 indicators from IMETA metadata

## Missing Data Statistics

```{r}
knitr::kable(imputation_report, caption = "Imputation Process Summary")
```

**Note:** `r imputation_report[imputation_report$Metric == "Total Missing Values", "Count"]` missing values across `r imputation_report[imputation_report$Metric == "Variables with Missing Data", "Count"]` of `r n_indicators_full` indicators, affecting `r imputation_report[imputation_report$Metric == "Countries with Missing Data", "Count"]` of `r n_countries` countries, were imputed prior to standardization.

# Descriptive Statistics Tables

## Descriptive Statistics - Original Data

The table below presents descriptive statistics for all Level 1 indicators in the original dataset before imputation:

```{r descriptives-gii-raw}
original_stats |>
  mutate(Indicator = indicator_names[Indicator]) |>
  knitr::kable(caption = paste0("Descriptive Statistics - Original Data (n=", n_indicators_full, " indicators)"), digits = 2)
```

## Descriptive Statistics - Complete Data

The table below presents descriptive statistics for all Level 1 indicators in the complete dataset after kNN imputation and case-wise cleaning:

```{r descriptives-gii-imputed}
complete_stats |>
  mutate(Indicator = indicator_names[Indicator]) |>
  knitr::kable(caption = paste0("Descriptive Statistics - Complete Data (n=", n_indicators_full, " indicators)"), digits = 2)
```